\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage[numbers,sort&compress]{natbib}
\usepackage{tikz}
\usetikzlibrary{arrows.meta,positioning,fit,backgrounds,shapes.geometric}
\usepackage{pgfplots}
\pgfplotsset{compat=1.17}

\title{CASPO: Learned Prefix Value Modeling for\\
       Stepwise Credit Assignment in Reasoning RL}
\author{Anonymous}
\date{}

\newcommand{\caspo}{\textsc{CASPO}}
\newcommand{\ppo}{PPO}
\newcommand{\ppocritic}{PPO+Critic}
\newcommand{\grpo}{GRPO}
\newcommand{\ipvrm}{IPVRM}
\newcommand{\vineppo}{VinePPO}

\begin{document}
\maketitle

\begin{abstract}
We propose \caspo{} (Credit Assignment Step Policy Optimization), a method for
reinforcement learning of mathematical reasoning that obtains step-level
credit signals from a learned prefix-value model rather than from on-policy
Monte Carlo prefix rollouts. \caspo{} retains the \vineppo{} step-segmentation
and step-TD advantage structure but replaces \vineppo{}'s $K{=}9$ rollouts at
each non-terminal prefix with a single forward pass through a learned
process reward model (PRM), trained to predict
$P(\text{success} \mid \text{prefix})$ from Monte Carlo step labels.
A periodic refresh of the PRM on rollouts of the current policy controls
distribution drift across RL training. The result is a strict drop-in for
\vineppo{} that uses the same PPO clipped objective, the same KL penalty,
the same response segmentation, and the same global PPO minibatch, but
eliminates the per-step Monte Carlo expansion that dominates \vineppo{}'s
wall-clock cost. We compare four methods (\grpo{}, \ppocritic{},
\vineppo{}, \caspo{}) under one trainer, one verifier, one rollout engine,
and one set of hyperparameters, on Qwen2.5-Math-1.5B with the
\texttt{dsr\_sub} subset of DeepScaleR (1209 prompts from One-Shot-RLVR),
with DeepSeekMath-7B on MATH-lighteval as a 7B paper-faithful reference. \ppocritic{} is the
canonical Schulman-2017 \ppo{} with a learned scalar value head on a
stripped-LM-head backbone copy and GAE; we use it rather than bare \ppo{}
(uniform sequence-broadcast advantages with no value baseline) because
\grpo{} already covers the value-free case via its per-prompt
group-relative baseline. \ppocritic{} provides a per-token learned baseline
that contrasts with \caspo{}'s per-step learned baseline and with
\vineppo{}'s on-policy Monte-Carlo step values. Throughout we keep the
algorithmic contract identical except for the source of the per-step value
$V(s_t)$, isolating the credit-signal change from confounders. Empirical
results are reported in Section~\ref{sec:results} after the corresponding
training and evaluation runs complete.
\end{abstract}

\section{Introduction}
\label{sec:intro}

Reinforcement learning has emerged as the dominant approach for improving the
mathematical reasoning of large language models, but the dominant signal is
sparse: a verifier returns 0/1 only at the end of a long chain of thought,
leaving the optimizer to assign credit across hundreds of intermediate tokens.
\ppo{} \citep{schulman2017ppo} addresses this by broadcasting a sequence-level
advantage uniformly across response tokens; \grpo{} \citep{deepseekmath2024}
refines the variance term using a group-relative baseline. Neither method
distinguishes a correct intermediate inference from a lucky guess that
nonetheless reached the right final answer.

\vineppo{} \citep{vineppo2024} addresses this gap by introducing
\emph{step-level} value estimates: the response is segmented into reasoning
steps, and the value at each step boundary $s_t$ is estimated by sampling $K$
on-policy continuations from $s_t$ and averaging their terminal verifier
rewards. This produces well-calibrated step values that bear out
empirically---\vineppo{} reports substantial improvements over \ppo{} and
\grpo{} on MATH and GSM8K---but at a cost: each non-terminal step boundary
requires $K$ additional generations, multiplying wall-clock time by roughly
the average step count per response.

This cost is most punishing precisely on the long-chain regimes where
step-level credit is most needed. On the One-Shot-RLVR \texttt{dsr\_sub}
DeepScaleR subset with Qwen2.5-Math-1.5B, the LaTeX-aware step splitter
produces ${\sim}40$ step boundaries per response (vs.\ ${\sim}10$--$20$ on
upstream MATH), and the $K{=}9$ Monte Carlo continuations push per-step
training to ${\sim}33$ minutes on $4 \times$H100 80GB. A 600-outer-step
\vineppo{} run at this configuration thus requires roughly $14$ days of
wall-clock and is practically infeasible to run as a baseline at our
experimental scale; the cost-effectiveness gap to \vineppo{} is largest
exactly where step-level credit assignment matters most.

\caspo{} replaces those Monte Carlo prefix rollouts with a single forward
pass through a \emph{learned} prefix value model $V_\phi(s_t)$, a process
reward model (PRM) trained to predict $P(\text{success} \mid s_t)$ from
Monte Carlo step labels collected on the SFT policy. \caspo{} keeps every
other \vineppo{} component fixed: the same LaTeX-aware step splitter, the
same step-TD advantage construction, the same clipped PPO objective, and
the same KL penalty. Step-level credit is therefore extracted from
$V_\phi$ in $\mathcal{O}(1)$ forward passes per response rather than
$\mathcal{O}(K \cdot \text{steps})$ generations, and a periodic
\emph{refresh} of $V_\phi$ on rollouts of the current policy controls the
distribution shift induced by RL.

\paragraph{Contributions.} We contribute:
(i) \caspo{}: a method that drops \vineppo{}'s prefix Monte Carlo step in
favor of a learned process reward model (PRM) trained to predict
$P(\text{success} \mid s_t)$ from Monte Carlo step labels, plugged into
the \vineppo{} step-TD advantage construction;
(ii) two step-TD advantage formulations of the prefix value: $\Delta p$
(default), the canonical match for a binary verifier, and $\Delta\log p$
(ablation), which operates on the log-probability scale of the underlying
token logits and amplifies rescue signal on the failure tail;
(iii) an iterated-refresh configuration that periodically retrains
$V_\phi$ from scratch on Monte-Carlo-labeled rollouts of the current
policy, converting the credit model from a single drifting model into a
sequence of in-distribution snapshots at amortized wall-clock cost well
below one \vineppo{} outer-step budget; we ablate refresh against the
no-refresh default that holds the Phase-1 $V_\phi$ fixed throughout RL;
(iv) a unified comparison stack that implements \grpo{}, \ppocritic{},
\vineppo{}, and \caspo{} under one trainer, one verifier, and one rollout
engine, controlling everything except the algorithmic credit-signal choice; and
(v) implementation infrastructure for full-finetune 7B-scale RL with
FSDP-sharded trainer and rank-local vLLM generation, including a streaming
CUDA-IPC weight-sync path that is collective with
\texttt{summon\_full\_params}, fp32 master weights with on-demand AdamW
CPU-offload during the sync window to fit the unsharded fp32 materialization,
and per-method ckpt sharding across multiple NVMe drives so each of the four
1B methods runs to completion in parallel without disk contention.

\paragraph{Implementation provenance.} The codebase is an adaptation and
extension of the public \vineppo{} setup \citep{vineppo2024,vineppocode}. The
DeepSeekMath-7B configuration's prompt template, rollout shape, PPO
hyperparameters, evaluation protocol, and LaTeX-aware step segmentation are
matched to or derived from the upstream paper and reference repository so
that any wall-clock or accuracy gap between methods reflects the algorithmic
change rather than infrastructure. The Qwen2.5-Math-1.5B configuration on
\texttt{dsr\_sub} adapts the same training contract to the model's native
prompt template and to a 4-GPU FSDP topology, and is the primary
experimental site for the comparison and for the iterated-refresh
contribution.

\section{Related Work}
\label{sec:related}

\paragraph{Policy gradient with sparse rewards.}
\ppo{} \citep{schulman2017ppo} optimizes a clipped policy-gradient surrogate
to bound the per-step policy update. Applied to language models with
verifier-only terminal rewards, the standard practice is to broadcast the
sequence-level advantage uniformly across response tokens and add a
KL-to-reference penalty to prevent the policy from drifting far from the SFT
initialization.

\paragraph{Group-relative baselines.}
\grpo{} \citep{deepseekmath2024} samples $G$ responses per prompt and uses
the within-group reward statistics as the advantage baseline. This avoids
training a separate value network entirely, at the cost of treating every
intermediate token in a correct response as equally important.

\paragraph{Step-level credit assignment via Monte Carlo prefix values.}
\vineppo{} \citep{vineppo2024} introduces stepwise value estimates by
sampling $K$ on-policy continuations from each non-terminal prefix and
averaging the verifier outcomes of those continuations. Step advantages are
then formed as temporal-difference differences between consecutive prefix
values. The reasoning-credit gain is well demonstrated on MATH, but the
method's rollout multiplier scales as $K$ times the average number of steps
per response.

\paragraph{Learned process supervision.}
A line of work trains a separate process reward model (PRM) on
human-annotated or self-supervised step-correctness labels. \caspo{}'s
PRM is in this family: it predicts $P(\text{success} \mid \text{prefix})$
from automatically-generated Monte Carlo step labels (a held-out batch
of $J$ rollouts is averaged to form the soft target at each prefix),
avoiding the human-annotation bottleneck of PRM800K-style data and
recovering a directly calibrated correctness probability suitable for
TD-style step advantages.

\paragraph{Implicit prefix value learning (\ipvrm{}).}
\ipvrm{} \citep{ipvrm2026} formalizes prefix values as cumulative
log-ratios of a trainable prefix policy versus a frozen reference policy,
trained from trajectory-level outcome labels using a margin BCE. We
\emph{adopt \ipvrm{}'s cumulative-log-ratio architecture} for $V_\phi$
and its framing as a calibrated correctness predictor that drives a
step-TD advantage. Our differences are in \emph{training-data form} and
\emph{refresh strategy}: we train $V_\phi$ on dense Monte Carlo step
labels with a continuous-target BCE (rather than trajectory-level
labels with the margin term), and we replace inline online updates
with periodic full retraining of $V_\phi$ from scratch on current-policy
rollouts. We also empirically verified that the cumulative-log-ratio
parameterization is essential: at matched hyperparameters and training
data, an alternative single-sigmoid-head parameterization
$V_\phi(s_t) = \sigma(W h_\phi(s_t))$ converged at val Spearman
$\rho \approx 0.43$ versus $\rho \approx 0.78$ for the
log-ratio architecture (matched step 1000), because the log-ratio's
$V_\phi(s_t) = \beta \sum_{i<t} \log\!\big(\pi_\phi/\pi_{\mathrm{ref}}\big)(a_i \mid s_{<i})$
decomposition yields one direct gradient term per response token rather
than a single end-of-prefix gradient.

\paragraph{Per-token learned-critic baselines.}
The classical PPO-with-critic recipe \citep{schulman2017ppo} uses a learned
scalar value head $V_\psi(s, t)$ trained jointly with the policy on the
value loss $(V_\psi - \hat R)^2$ and folded into the advantage via
generalized advantage estimation (GAE). We include this recipe as a baseline
(\ppocritic{}) so that \caspo{}'s learned \emph{step-level} prefix value can
be compared not only against the value-free baselines (\ppo{}, \grpo{}) and
the on-policy Monte Carlo step value (\vineppo{}), but also against an
in-line per-token learned baseline trained from the same trajectories under
the same PPO objective.

\paragraph{Generation backends and weight sync.}
We use vLLM \citep{kwon2023vllm} for high-throughput rollout and prefix
caching. The implementation uses CUDA-IPC weight transfer between trainer
ranks and rank-local vLLM engines; for the FSDP-sharded 7B trainer, the IPC
sync is implemented as a collective \texttt{summon\_full\_params} block over
the FSDP-wrapped policy followed by streaming per-tensor IPC handle exchange
with each rank's vLLM \texttt{EngineCore}, in chunks of 8 parameters with
per-parameter bf16 casting to bound transient host memory during the sync
window.

\section{Background}
\label{sec:bg}

\subsection{Reasoning RL setup}
A trajectory consists of a problem prompt $x$ and a model response
$y = (y_1, \dots, y_T)$ generated autoregressively. A verifier
$R(x,y) \in \{0,1\}$ returns 1 if and only if the boxed final answer in $y$
matches the ground truth. The training objective is to maximize
$\mathbb{E}_{x \sim \mathcal{D}, y \sim \pi_\theta(\cdot \mid x)} R(x,y)$
subject to a KL constraint to a frozen reference policy $\pi_{\mathrm{ref}}$,
which we instantiate as the SFT initialization.

\subsection{Step segmentation}
We split each response into reasoning steps using the LaTeX-aware splitter of
\vineppo{} \citep{vineppo2024,vineppocode}, which identifies natural
mathematical step boundaries (display equations, aligned blocks, sentence
terminators in prose). Let $0 = t_0 < t_1 < \dots < t_S = T$ be the resulting
boundary token indices and let $s_t \triangleq y_{<t}$ denote the prefix
state at boundary $t$.

\subsection{Step-TD advantage}
Given a per-step value function $V(s_t)$, both \vineppo{} and \caspo{}
construct the step-TD advantage
\begin{equation}
\label{eq:adv}
A(s_t) = r_t + \gamma\, V(s_{t+1}) - V(s_t),
\qquad
r_t = \begin{cases} R(x,y) & t = S, \\ 0 & t < S, \end{cases}
\end{equation}
with $\gamma = 1$. \caspo{} obtains $V$ from a learned model
$V_\phi$; \vineppo{} obtains $V$ by averaging $K$ Monte Carlo continuations.
The downstream PPO objective and KL penalty are identical for both methods;
only the source of $V$ differs.

\subsection{PPO objective}
We use the clipped PPO objective
\begin{equation}
\label{eq:ppo}
\mathcal{L}_{\mathrm{clip}}(\theta) =
-\,\mathbb{E}_t\!\left[
\min\!\big(
\rho_t(\theta)\, A_t,\;
\mathrm{clip}\!\big(\rho_t(\theta), 1-\epsilon, 1+\epsilon\big)\, A_t
\big)
\right],
\end{equation}
with $\rho_t(\theta) = \pi_\theta(y_t \mid x, y_{<t}) / \pi_{\theta_{\mathrm{old}}}(y_t \mid x, y_{<t})$
and $\epsilon = 0.2$. A KL-to-reference term is added with the $k_3$
estimator:
\begin{equation}
\label{eq:kl}
\mathrm{KL}_{k_3}(\pi_\theta \,\Vert\, \pi_{\mathrm{ref}}) \approx
\mathbb{E}_t\!\left[
\exp(\delta_t) - \delta_t - 1
\right], \quad
\delta_t = \log \pi_{\mathrm{ref}}(y_t) - \log \pi_\theta(y_t).
\end{equation}
The advantages $A_t$ in \eqref{eq:ppo} are (a) terminal-broadcast group-relative
advantages for \grpo{}; (b) GAE per-token advantages from a learned scalar
value head for \ppocritic{}; or (c) the step-TD advantages of \eqref{eq:adv}
broadcast over the tokens in each step span for \vineppo{}/\caspo{}.

\section{Method: \caspo{}}
\label{sec:method}

\begin{figure}[t]
\centering
\begin{tikzpicture}[
    font=\footnotesize,
    node distance=4mm,
    box/.style={rectangle, rounded corners=2pt, draw, thick, minimum width=20mm, minimum height=8mm, align=center, inner sep=2pt},
    src/.style={box, fill=black!8},
    valbox/.style={box, fill=black!4, minimum width=33mm},
    arr/.style={->, >=Stealth, thick},
    grp/.style={rectangle, draw, dashed, rounded corners=3pt, inner sep=4mm},
]
% Shared rollout column on the left
\node[src] (resp) {response $y$\\(rollout)};
\node[src, below=of resp] (split) {LaTeX-aware\\step splitter};
\node[src, below=of split] (verif) {verifier\\$R(x,y)\!\in\!\{0,1\}$};
\draw[arr] (resp) -- (split);
\draw[arr] (split) -- (verif);

% Four method columns, anchored on the value-source step
\node[valbox, right=12mm of resp.east] (vgrpo)
    {\grpo{}\\
     $V(s_t)\!\equiv\!0$\\
     group-relative terminal\\broadcast};
\node[valbox, below=of vgrpo] (vpc)
    {\ppocritic{}\\
     $V_\psi(s,t)$ scalar head\\
     per-token GAE};
\node[valbox, below=of vpc] (vvine)
    {\vineppo{}\\
     $K\!=\!9$ MC continuations\\
     $V(s_t)=\bar R_K$};
\node[valbox, below=of vvine] (vcaspo)
    {\caspo{}\\
     $V_\phi(s_t) \approx P(\text{success}\mid s_t)$\\
     learned PRM (\ipvrm{} log-ratio)};

% Step-TD / advantage column
\node[box, right=18mm of vpc.east, minimum width=33mm, fill=black!4] (adv)
    {step-TD advantage\\$A_t = r_t + \gamma V(s_{t\!+\!1}) - V(s_t)$};
\node[box, below=of adv, minimum width=33mm, fill=black!4] (clip)
    {PPO clipped objective\\$+ k_3$ KL penalty};

% Arrows from each method to the advantage box
\draw[arr] (vgrpo.east) -- (adv.west);
\draw[arr] (vpc.east) -- (adv.west);
\draw[arr] (vvine.east) -- (adv.west);
\draw[arr] (vcaspo.east) -- (adv.west);
\draw[arr] (adv) -- (clip);

% Source-of-value annotation bracket
\node[draw=none, fit=(vgrpo)(vcaspo), label={[anchor=south]above:\textbf{value source} $V(s_t)$}] {};
\end{tikzpicture}
\caption{Four compared methods, anchored on the source of the per-step
value $V(s_t)$. Rollout, segmenter, verifier, advantage form, PPO clipped
objective, and KL penalty are identical across all rows; the only difference
is which black box produces $V(s_t)$ used in
$A_t = r_t + \gamma V(s_{t\!+\!1}) - V(s_t)$. \grpo{} uses no value
($V \equiv 0$) with a per-prompt group-relative terminal broadcast.
\ppocritic{} (the canonical Schulman-2017 \ppo{}) learns a scalar per-token
value head trained on $(V_\psi - \hat R)^2$ with GAE. \vineppo{} estimates
$V(s_t)$ by averaging $K{=}9$ on-policy continuations from each prefix.
\caspo{} replaces those continuations with a learned process reward
model $V_\phi(s_t)$ trained on Monte Carlo step labels to approximate
$P(\text{success}\mid s_t)$, with periodic refresh on current-policy
rollouts (§\ref{sec:vphi}, §\ref{sec:refresh}).}
\label{fig:methods}
\end{figure}

Figure~\ref{fig:methods} summarizes the comparison axis: rollout, step
splitter, verifier, advantage form, PPO objective, and KL penalty are held
identical across all four methods. Only the source of the per-step value
$V(s_t)$ differs.

\subsection{Prefix value model}
\label{sec:vphi}
The prefix value model $V_\phi$ is a neural network trained to predict
the probability that a rollout starting from prefix $s_t$ eventually
reaches a correct final answer:
\begin{equation}
\label{eq:vphi}
V_\phi(s_t) \;\approx\; P\big(R(x, y) = 1 \mid s_t\big).
\end{equation}
Following \ipvrm{} \citep{ipvrm2026} we parameterize $V_\phi$ as a
\emph{cumulative log-ratio} of a trainable prefix policy
$\pi_\phi$ against a frozen reference policy $\pi_{\mathrm{ref}}$
(both initialized from the same SFT checkpoint as the base policy):
\begin{equation}
\label{eq:vphi_param}
V_\phi(s_t) \;=\; \beta \sum_{i<t}
   \log\!\frac{\pi_\phi(a_i \mid s_{<i})}{\pi_{\mathrm{ref}}(a_i \mid s_{<i})},
\end{equation}
with $\beta = 10$ a fixed stability scale. The PRM probability is
recovered by a sigmoid on the scaled logit
$\sigma\!\big(V_\phi(s_t)/\beta\big) \in [0, 1]$. At initialization
$\pi_\phi = \pi_{\mathrm{ref}}$, so $V_\phi \equiv 0$ everywhere and
the implied probability is exactly $0.5$ for every prefix --- a
natural calibrated prior. We refer to $V_\phi$ as the \emph{process
reward model} (PRM), following the convention of \citet{ipvrm2026}
and the process-supervision literature. The cumulative-log-ratio
form is critical: because the value at any prefix decomposes as a
sum of per-token log-ratio terms, every response token contributes a
direct gradient signal during training rather than only the
single-token signal an end-of-prefix scalar head would provide. We
verified empirically that replacing this parameterization with
$V_\phi(s_t) = \sigma(W h_\phi(s_t))$ at matched hyperparameters
collapses validation Spearman $\rho$ from $\approx 0.78$ to
$\approx 0.43$ (\S\ref{sec:phase1}).

\subsection{Phase-1: PRM training via Monte Carlo step labels}
\label{sec:phase1}
We train $V_\phi$ directly on Monte Carlo step labels rather than via a
trajectory-level BCE-with-margin objective. For each of $N$ training
prompts we sample $K$ base rollouts from the SFT policy at temperature
$1.0$. From each rollout we randomly select $S$ step boundaries; at each
boundary $s_t$ we sample $J$ additional independent continuations from
the same SFT policy, each scored by the verifier $R \in \{0, 1\}$, and
form the empirical Monte Carlo estimate
\begin{equation}
\label{eq:phat}
\widehat{p}(s_t) \;=\; \frac{1}{J}\sum_{j=1}^{J} R\!\big(x, [s_t; y^{(j)}_{>t}]\big),
\qquad y^{(j)}_{>t}\sim \pi_{\mathrm{SFT}}(\cdot \mid s_t),
\end{equation}
which is an unbiased estimator of $P(R = 1 \mid s_t)$. Following the
process-supervision practice of mixed-outcome filtering, we discard
rollouts whose $J$ continuations are all 0 or all 1 at every sampled
boundary, since saturated prefixes contribute no learning signal.

We train $V_\phi$ to minimize binary cross-entropy against the
continuous targets $\widehat{p}(s_t) \in [0, 1]$, applied to the
sigmoid of the scaled cumulative log-ratio:
\begin{equation}
\label{eq:prmloss}
\mathcal{L}_{\mathrm{PRM}}(\phi) \;=\;
-\,\mathbb{E}_{s_t}\!\Big[\,
\widehat{p}(s_t)\,\log \sigma\!\big(V_\phi(s_t)/\beta\big)
\;+\;
\big(1-\widehat{p}(s_t)\big)\,\log\!\big(1 - \sigma\!\big(V_\phi(s_t)/\beta\big)\big)
\Big],
\end{equation}
with $V_\phi(s_t)$ defined by the cumulative log-ratio
of \eqref{eq:vphi_param}. Equation~\eqref{eq:prmloss} is BCE on
continuous targets in $[0, 1]$ rather than binary; it minimizes the
Bregman divergence between the implied PRM probability
$\sigma\!\big(V_\phi(s_t)/\beta\big)$ and the MC label
$\widehat{p}(s_t)$, and recovers the same calibration target without
needing the trajectory-level margin term of \ipvrm{}
\citep{ipvrm2026}. Concretely, the trainer caches the per-token
log-probabilities of $\pi_\phi$ and $\pi_{\mathrm{ref}}$ on the
prefix, forms the per-token log-ratios, takes a cumulative sum, and
gathers the value at the step boundary $s_t$. Because all $i \le t$
log-ratio terms enter the loss additively, every response token in
the prefix produces a direct gradient signal --- which is the
empirical reason this parameterization dominates an end-of-prefix
scalar-head alternative (\S\ref{sec:related}). We train for two
epochs with full-model FSDP updates, $\beta = 10$, effective batch
size 32, and learning rate $5 \times 10^{-6}$.

\paragraph{Held-out PRM evaluation.}
For PRM quality reporting we evaluate $V_\phi$ on a \emph{prompt-level}
held-out split rather than a prefix-level shuffle. The default
\texttt{train\_value\_mc.py} flow exposes two non-leaky modes:
\texttt{--split\_by\_prompt}, which holds out
$\lceil \mathrm{val\_fraction} \cdot |\mathcal{D}_{\mathrm{prompt}}|\rceil$
unique prompts (and all their prefixes) from the same N=300 collection,
and \texttt{--held\_out\_data <path>}, which loads a separately-collected
mc\_labels file generated on disjoint prompts (the remaining 909
\texttt{dsr\_sub} prompts not in the primary training collection). The
prefix-level row shuffle is preserved as a legacy default but flagged
explicitly as leaky in the trainer's startup banner.

\subsection{Phase-2: step-TD policy optimization}
At each PPO outer step, the trainer samples $G$ responses per prompt
for $P$ prompts, segments each response, and constructs step values
$\{V_\phi(s_t)\}$ in a single forward pass through $V_\phi$ using
\eqref{eq:vphi}. Step advantages are assembled from \eqref{eq:adv} and
broadcast over the tokens within each step span before applying the
PPO objective in \eqref{eq:ppo} with the KL penalty in \eqref{eq:kl}.
The default \caspo{} run holds $V_\phi$ frozen during policy
optimization; an iterated-refresh variant (next section) periodically
retrains $V_\phi$ from scratch on rollouts of the current policy to
control distribution drift.

\subsection{Iterated PRM refresh}
\label{sec:refresh}

\paragraph{PRM decay.}
A learned PRM is in-distribution only at the policy state on which its
MC labels were collected. As RL training updates the policy, the
prefixes seen at outer step $t$ drift away from the policy that
generated the Phase-1 labels, and $V_\phi$'s calibration on those
prefixes deteriorates. Concretely, on Qwen2.5-Math-1.5B with the
\texttt{dsr\_sub} dataset we
measure the per-checkpoint Spearman $\rho$ between $V_\phi(s_t)$ and a
fresh $J{=}8$ MC estimate of $P(\text{success} \mid s_t)$ at policy
states $\{$step\_50, step\_100, $\ldots$, step\_400$\}$. A PRM trained
on rollouts of step\_0 (SFT) yields $\rho \approx 0.63$ when scored on
its own distribution; on policy state step\_150 this drops to
$\rho \approx 0.54$, and on step\_400 it falls to $\rho \approx 0.27$.
The decay is monotone with cumulative policy drift and is the dominant
mechanism by which a fixed prefix-value model loses its credit-signal
quality during RL.

\paragraph{Refresh mechanism.}
We address PRM decay by replacing a single drifting $V_\phi$ with a
sequence of in-distribution snapshots, refreshed periodically. Every
$k$ outer steps the trainer pauses, collects a new MC label set from
the current policy snapshot under the same Phase-1 recipe
(\eqref{eq:phat}, \eqref{eq:prmloss}), retrains $V_\phi$ \emph{from
scratch} on those labels (initialized from the SFT backbone, not from
the previous $V_\phi$, to avoid bias compounding across refreshes),
and resumes RL using the new $V_\phi$ as the prefix value source. We
use $k = 150$ for the $\Delta p$ formulation and $k = 200$ for
$\Delta\log p$, chosen so that $\rho$ at the refresh trigger remains
well above the empirical $\tau \approx 0.4$ floor at which the
credit-signal contribution of $V_\phi$ becomes negligible.

\paragraph{Refresh cost.}
A single refresh on Qwen2.5-Math-1.5B with \texttt{dsr\_sub} takes
$\approx 91$ minutes on $4 \times$H100 80GB ($\approx 41$~min for MC
collection at $K{=}16$, $J{=}8$ on $N{=}300$ prompts, plus
$\approx 50$~min for FSDP=4 PRM training). At $k = 150$ this amortizes
to $\approx 36$~s of refresh overhead per RL outer step, raising
total \caspo{}+refresh wall-clock for a 600-step RL run by roughly
$5{-}6$~h above the no-refresh \caspo{} baseline. The amortized
refresh cost is small relative to overall per-step training time and
is an order of magnitude below \vineppo{}~$K{=}9$.

\paragraph{Ablation.}
We report \caspo{} both with and without refresh: \caspo{} (the
no-refresh default holds the Phase-1 $V_\phi$ fixed throughout RL)
and \caspo{}+refresh (the schedule above). The two variants share
every other component (PPO objective, KL, advantage transform,
segmenter), isolating the refresh contribution to a single ablation.

\subsection{Step-TD advantage formulations from $V_\phi$}
\label{sec:transforms}

A modeling choice in \caspo{} is \emph{which} statistic of $V_\phi$
enters the step-TD difference. Because $V_\phi(s_t) \in [0, 1]$ is
already a calibrated correctness probability under \eqref{eq:vphi},
two natural formulations arise: the TD on probabilities ($\Delta p$,
default) and the TD on log-probabilities ($\Delta\log p$, ablation).

\begin{description}
\item[$\Delta p$] (default, probability-scale TD):
\begin{equation}
\label{eq:advP}
A^{\Delta p}(s_t) \;=\; r_t \,+\, \gamma\, V_\phi(s_{t+1}) \,-\, V_\phi(s_t).
\end{equation}
\item[$\Delta\log p$] (ablation, log-probability-scale TD):
\begin{equation}
\label{eq:advLP}
A^{\Delta\log p}(s_t) \;=\; r_t \,+\, \gamma\, \log V_\phi(s_{t+1}) \,-\, \log V_\phi(s_t).
\end{equation}
\end{description}

In both variants the terminal verifier reward $r_T = R(x,y)$ enters
identically; the only difference is whether the TD operates on
$V_\phi$ or $\log V_\phi$.

\paragraph{Intuition.}
\begin{itemize}
\item $\Delta p$: The TD operates on the probability scale, so
$|A^{\Delta p}(s_t)| \leq 1$ before standardization. Because
$V_\phi(s_t) \approx P(R = 1 \mid s_t)$, $A^{\Delta p}$ is the
empirical advantage an idealized binary-reward critic would compute:
the change in expected terminal reward attributed to step $t$. This
matches the binary verifier exactly. Its main weakness is
\emph{near-saturation}: when $V_\phi(s_t)$ is close to $0$ or $1$,
$\Delta p$ contributions are vanishingly small, so highly confident
prefixes contribute little gradient.

\item $\Delta\log p$: The TD operates on the log-probability scale,
mirroring the form of token log-likelihoods used in the PPO ratio
$\rho_t(\theta)$. For a sequence model trained by maximum likelihood,
log-probability differences are the canonical update signal; using
$\log V_\phi$ keeps the value-derived advantage on the same
information-theoretic scale as the underlying token logits. The
log-transform also amplifies the \emph{rescue} signal: a step that
moves a near-failed trajectory ($V_\phi \to 0$) toward feasibility
contributes a large negative-side gradient under $\log V_\phi$ but
only a small positive-side gradient under $V_\phi$. The trade-off is
unbounded magnitude on the failure tail, which we control with the
shared $\pm 3$ standardized advantage clip in
Section~\ref{sec:method}.
\end{itemize}

\paragraph{When each formulation is preferable.}
$\Delta p$ is the default: it is the canonical match for a binary
verifier and has the clean expected-terminal-reward interpretation
$A^{\Delta p}(s_t) = r_t + \gamma\, \mathbb{E}[R \mid s_{t+1}] -
\mathbb{E}[R \mid s_t]$ via \eqref{eq:vphi}. $\Delta\log p$ is an
ablation that exchanges saturation-near-1 robustness for
unbounded-magnitude gradients on the failure tail; we report it as
the second formulation. We report both on MATH-500 in
Section~\ref{sec:results}.

\begin{figure}[t]
\centering
\begin{tikzpicture}
\begin{axis}[
    width=\linewidth, height=58mm,
    xlabel={prefix value $V_\phi(s_t) \in [0,1]$},
    ylabel={transformed value},
    xmin=0, xmax=1, ymin=-5.5, ymax=1.0,
    grid=both,
    grid style={black!12},
    legend style={font=\footnotesize, at={(0.02,0.98)}, anchor=north west, draw=black!50},
    axis line style={->},
    xtick={0,0.2,0.4,0.6,0.8,1.0},
    ytick={-5,-4,-3,-2,-1,0,1},
    samples=121,
    domain=0.005:0.995,
]
\addplot[thick]                        {x};                  \addlegendentry{$\Delta p$: $V_\phi$}
\addplot[thick, dotted, line width=1pt] {ln(x)};             \addlegendentry{$\Delta\log p$: $\log V_\phi$}
\addplot[thin, gray] coordinates {(0,0)(1,0)};
\end{axis}
\end{tikzpicture}
\caption{The two step-TD transforms of the prefix value $V_\phi(s_t)
\in [0,1]$. $\Delta p$ (solid) is the identity: bounded magnitude
$|A^{\Delta p}| \leq 1$ before standardization, but saturates near
$V_\phi = 0$ and $V_\phi = 1$ where two prefixes the model is
confident about contribute essentially the same gradient.
$\Delta\log p$ (dotted) is unbounded as $V_\phi \to 0$, amplifying
\emph{rescue} signal on the failure tail (where most early-training
gradient lives) and compressing the high-confidence regime. Both
transforms share the global advantage clip $|A| \leq 3$ in
Section~\ref{sec:method}. The TD difference
$A^*(s_t) = r_t + \gamma\,f(V_\phi(s_{t+1})) - f(V_\phi(s_t))$
inherits the regime of $f$.}
\label{fig:transforms}
\end{figure}

\subsection{Step advantage normalization and clipping}
Let $\mathcal{A} = \{A(s_t)\}$ over all valid step boundaries in the outer
PPO step. We standardize over the entire batch
($\widetilde{A}_t = (A_t - \mu_{\mathcal{A}}) / (\sigma_{\mathcal{A}} +
\varepsilon)$) and clip the standardized advantages to $[-3, 3]$ to bound the
contribution of single-step outliers. The clip range is chosen empirically to
balance suppression of pathological prefixes against retaining genuine
step-level signal.

\section{Experimental Setup}
\label{sec:setup}

\subsection{Models and data}
The primary experimental site is \textbf{Qwen2.5-Math-1.5B}
(\texttt{Qwen/Qwen2.5-Math-1.5B}), a 1.5B math-specialist SFT
initialization, trained on the \textbf{One-Shot-RLVR \texttt{dsr\_sub}}
1209-prompt subset of DeepScaleR. Evaluation uses MATH-500, GSM8K, and
OlympiadBench under greedy decoding ($k{=}1$, $T{=}0$). Qwen2.5-Math is a
math-specialist SFT model that responds best to its native template
\texttt{\{query\}\textbackslash nLet's think step by step and output the
final answer within \textbackslash boxed\{\}.}; substituting the
\texttt{[MATH\_TASK]} template degrades Qwen accuracy by ${>}30$ percentage
points and is a critical reproduction footgun.

For a 7B paper-faithful reference site, we additionally evaluate on
\textbf{DeepSeekMath-7B} (\texttt{realtreetune/deepseekmath-7b-sft-MATH-v2}),
the SFT-on-MATH initialization from the upstream \vineppo{} pipeline,
trained on \texttt{DigitalLearningGmbH/MATH-lighteval} (train split). 7B
held-out evaluation uses \texttt{HuggingFaceH4/MATH-500} (test split, 500
problems) plus the full MATH test set, College Math, and OlympiadBench
(math-only configuration), under the upstream prompt template
\texttt{[MATH\_TASK] Problem:\textbackslash n\{query\}\textbackslash n\textbackslash nSolution:}
and the LaTeX-aware step splitter from the upstream codebase.

\subsection{Methods}
We compare four methods (Figure~\ref{fig:methods}):
\begin{description}
\item[\grpo{}] Per-prompt group-relative terminal-reward advantages with
$G=8$. Stands in for the value-free baseline; bare \ppo{} (uniform
sequence-broadcast advantages, no group baseline, no value head) is
strictly weaker on this evaluation regime and is omitted.
\item[\ppocritic{}] Schulman-2017 \ppo{} with a learned scalar value head
$V_\psi$ on a stripped-LM-head backbone copy, trained jointly with the
policy under a clipped value loss
$\max\!\big[(V_\psi - \hat R)^2,\, (\mathrm{clip}(V_\psi, V_\psi^{\mathrm{old}} \pm \epsilon_v) - \hat R)^2\big]$
with $\epsilon_v = 0.2$ and $\lambda_{\mathrm{GAE}} = 1.0$, value-loss
coefficient $1.0$, critic LR $10^{-6}$.
\item[\vineppo{}] $K=9$ Monte Carlo continuations from each non-terminal step
boundary; step-TD advantages.
\item[\caspo{}] Learned process reward model
$V_\phi(s_t) \approx P(\text{success} \mid s_t)$, an \ipvrm{}-style
cumulative-log-ratio network trained on Monte Carlo step labels
(Section~\ref{sec:phase1}), plugged into the step-TD advantage. We report two step-TD formulations
(Section~\ref{sec:transforms}): \caspo{}-$\Delta p$ (default) and
\caspo{}-$\Delta\log p$ (ablation). Each is reported with and without
the iterated PRM-refresh schedule of Section~\ref{sec:refresh},
labeled \caspo{} and \caspo{}+refresh respectively, where the
no-refresh default holds the Phase-1 $V_\phi$ fixed throughout RL.
\end{description}

\subsection{Hyperparameters}
All four methods share the same PPO infrastructure. Table~\ref{tab:hp}
summarizes the key knobs.

\begin{table}[t]
\centering
\caption{Training hyperparameters at the two experimental sites. The
Qwen2.5-Math-1.5B configuration is the primary site for the comparison;
DeepSeekMath-7B retains the upstream \vineppo{} configuration as a
paper-faithful 7B reference. The 2048-token response budget for
Qwen2.5-Math-1.5B is justified empirically (see text below).}
\label{tab:hp}
\begin{tabular}{lll}
\toprule
Field & Qwen2.5-Math-1.5B & DeepSeekMath-7B \\
\midrule
Group size $G$ & 8 & 8 \\
Prompts per outer step & 128 & 64 \\
Responses per outer step & 1024 & 512 \\
Global PPO minibatch & 128 & 64 \\
PPO epochs per rollout & 2 (\grpo{} reported at both $\mu{=}1$ and $\mu{=}2$\textsuperscript{a}) & 2 \\
Policy LR & $1\times 10^{-6}$ & $1\times 10^{-6}$ \\
Online value LR (\caspo{}) & $1\times 10^{-6}$ & $1\times 10^{-6}$ \\
Critic LR (\ppocritic{}) & $1\times 10^{-6}$ & $1\times 10^{-6}$ \\
Value $\beta$, $m$ & $10$, $5$ & $10$, $5$ \\
PPO clip $\epsilon$ & $0.2$ & $0.2$ \\
Value clip $\epsilon_v$ (\ppocritic{}) & $0.2$ & $0.2$ \\
GAE $\lambda$ (\ppocritic{}) & $1.0$ & $1.0$ \\
KL coefficient & $10^{-3}$ ($k_3$, $10^{-2}$ for \ppocritic{}) & $10^{-2}$ ($k_3$) \\
Advantage clip & $3.0$ & $3.0$ \\
Advantage standardization & batch & batch \\
Rollout sampling & temp $0.6$, top-$p$ $1.0$ & temp $0.6$, top-$p$ $0.9$ \\
Max prompt / response & 1024 / 2048 & 1512 / 1024 \\
Max sequence length & 3072 & 2499 \\
Warmup steps & 0 & 480 \\
Outer steps & 600 & 1000 \\
Checkpoint cadence & every 50 & every 200 \\
\bottomrule
\end{tabular}

\smallskip
\footnotesize
\textsuperscript{a} \ppocritic{}, \vineppo{}, and \caspo{} are fixed at
$\mu{=}2$ epochs per rollout, matching the upstream \vineppo{}
configuration. \grpo{} is reported at both $\mu{=}1$ (canonical setting
in DeepSeekMath \citep{deepseekmath2024}, TRL, and verl) and $\mu{=}2$
(matches the optimizer-step budget per rollout of the other three
methods, following the \vineppo{}-paper \grpo{} baseline). The $\mu{=}1$
comparison is \grpo{}-faithful; the $\mu{=}2$ comparison isolates the
credit-signal change from optimizer-budget effects.
\end{table}

The Qwen2.5-Math-1.5B configuration uses 1024 responses per outer step
(128 prompts $\times G{=}8$) on a 4-GPU FSDP topology. The KL coefficient
of $10^{-3}$ for \caspo{}/\grpo{}/\vineppo{} (versus $10^{-2}$ for
\ppocritic{}) is empirically required: the math-specialist Qwen SFT
initialization is over-anchored at $10^{-2}$ and policy updates do not
move the model without a weaker reference anchor, while \ppocritic{} is
critic-stabilized and tolerates the stronger anchor. The DeepSeekMath-7B
configuration retains the upstream \vineppo{} hyperparameters at 7B
\citep{vineppo2024,vineppocode}: 512 responses per outer step
($64 \times G{=}8$), 2 PPO epochs per rollout, and a 64-response global
PPO minibatch, isolating any empirical differences between the two sites
to the model and dataset rather than the optimization recipe.

The response budget of 2048 tokens for Qwen2.5-Math-1.5B is set by
empirical measurement on \texttt{dsr\_sub}: an 800-chain uncapped sample
($T{=}1.0$, max=3000) places the 98th percentile of \emph{correct} chains
at 1613 tokens. At $2048$, only $2\%$ of correct chains are truncated,
while the overall truncation rate is $8\%$ (i.e., the truncated tail is
overwhelmingly off-track / rambling chains rather than length-bound
correct ones). This makes the unfinished-response penalty signal load
the right pattern: ``long without resolution'' is what gets penalized,
not ``hard problem requires more tokens.'' Going further to $3072$ catches
fewer than one additional correct chain per 150 sampled and risks the
Qwen \texttt{max\_position\_embeddings}${=}4096$ ceiling at
$\text{prompt}+\text{response}$.

\subsection{Evaluation protocol}
Final evaluation reports pass@1 and pass@$k$ ($k=16$) on MATH-500, full MATH
test, College Math, and OlympiadBench. Sampling at evaluation uses
temperature $0.35$ and top-$p$ $0.9$, matching the upstream protocol.
Generation is performed with vLLM at \texttt{kv\_cache\_dtype="fp8"} and
\texttt{gpu\_memory\_utilization=0.92} for maximum eval throughput on a
dedicated GPU. Sample evaluation at intermediate checkpoints uses
\texttt{EVAL\_LIMIT=100}, \texttt{EVAL\_K=8} for low-overhead training-curve
visibility; the full benchmark suite is run at the final checkpoint.

\subsection{Infrastructure}
\label{sec:infra}

\paragraph{Trainer + colocated rollout.}
Both scales share a single PyTorch trainer plus rank-local vLLM rollout. The
1B configuration uses a single H100 per method (Figure~\ref{fig:sysarch},
left); the 7B configuration uses FSDP=4 (full\_shard, use\_orig\_params) for
\grpo{}/\ppocritic{}/\caspo{} and FSDP=8 for \vineppo{}, with vLLM
colocated rank-local in both layouts (Figure~\ref{fig:sysarch}, right).

\paragraph{Streaming CUDA-IPC weight sync.}
At 1B with a single-rank trainer, the IPC handle exchange is direct. At 7B
under FSDP, the trainer enters
\texttt{FSDP.summon\_full\_params(writeback=False)} collectively, then
streams parameters to each rank's vLLM
\texttt{EngineCore.collective\_rpc("update\_weights", \dots)} in chunks of 8
tensors with per-parameter bf16 cast inside the IPC push. Streaming bounds
peak host-side staging memory and avoids materializing the entire
unsharded fp32 policy on host before transfer.

\paragraph{fp32 master weights with on-demand AdamW offload.}
Both methods co-train policy (and \caspo{}'s value model, or \ppocritic{}'s
critic) in bf16 compute with fp32 master weights. Under FSDP, the master
copy lives in the AdamW optimizer state and is materialized into bf16
shards through \texttt{MixedPrecision(param\_dtype=bf16,
reduce\_dtype=fp32)}. At sync time the
\texttt{summon\_full\_params} block briefly materializes a $\sim$28~GiB
unsharded fp32 buffer per rank for the 7B policy. To fit alongside the
already-resident $\sim$14~GiB of AdamW state and the
$\sim$24~GiB vLLM KV cache reservation per H100, the trainer offloads the
AdamW state to pinned host memory before entering the summon block and
restores it after IPC push, costing $\sim$400--500\,ms per sync at PCIe Gen5
but freeing the GPU memory required for the unsharded fp32 materialization.

\paragraph{1B fp32 master via autocast.}
At 1B (single-GPU, no FSDP) the trainer wraps each forward in
\texttt{torch.autocast(dtype=bfloat16, cache\_enabled=False)} so the fp32
parameters cast to bf16 for compute on every op. The non-cached cast is
required for AdamW optimizer-state coverage to include all parameters; with
caching enabled, the autocast layer holds bf16 tensor wrappers that are not
written through to the underlying fp32 parameters during the optimizer step.

\paragraph{Multi-drive checkpoint sharding for parallel 1B training.}
Each method's 1B run produces $\sim$15--29~GB per checkpoint
(model in bf16, plus the AdamW master+momentum bundles for policy, plus
either \caspo{}'s value-model AdamW or \ppocritic{}'s critic AdamW). At
$\mathtt{save\_every}{=}100$ over 1000 outer steps that is roughly
150--290~GB per method-run. To run all four 1B method-runs (\grpo{},
\ppocritic{}, \caspo{}, \caspo{}-\emph{prob}) concurrently on four GPUs of
the same node at the $\mathtt{save\_every}{=}100$ cadence, each method's
outputs are routed to a dedicated 350~GB NVMe volume, so the four runs
neither compete for disk bandwidth nor cross-fill each other's drives.

\paragraph{Reference-policy sharing.}
The frozen reference policy is shared between the KL term and \caspo{}'s
value-model ref to avoid a duplicate forward. Rollout, old-logprob rescore,
ref-logprob precompute, and policy/value forward all use FlashAttention-3 on
Hopper.

\paragraph{Paper-faithful alignment with upstream \vineppo{}.}
A targeted audit against the upstream \vineppo{} reference repository
\citep{vineppocode} produced a set of small but consequential alignment
fixes that we include in the comparison stack so that our \vineppo{}
baseline reproduces the upstream numerics:
prepending the BOS token at both rollout and training,
per-microbatch advantage whitening,
adding the prompt-style stop string \texttt{"\textbackslash n\textbackslash n\textbackslash nProblem:"} for vLLM,
clamping per-token KL into $[0,10]$ before reduction,
penalizing unfinished sequences via the max-sequence-length policy in
\eqref{eq:ppo}, an explicit PPO ratio safety skip at $10\times$,
and uniform grad-accumulation per outer step. With these in place,
the \vineppo{} baseline in our stack matches the upstream
configuration field-for-field at the respective scale.

\paragraph{Stabilization deviations from upstream
(\ppocritic{}/\caspo{} only).}
Two deliberate departures from upstream-faithful numerics were required
to keep the online-value methods (\ppocritic{}, \caspo{}, \caspo{}-Δp)
stable at 1B. Empirically, with upstream's $\mathrm{kl\_coef}=10^{-4}$
and no advantage clip, both \ppocritic{} and \caspo{} drift to KL$\to
\infty$ within $\sim 20$ outer steps in our reimplementation; \grpo{}
remains stable. The fix was a stronger ref-anchor and an outlier safety
net:
\begin{itemize}
\item $\mathrm{kl\_coef} = 10^{-2}$ (100$\times$ upstream's $10^{-4}$,
  midway between InstructGPT's $0.02$ and DeepSeek-Math GRPO's $0.04$).
  This is roughly the standard RLHF range; upstream \vineppo{}'s unusually
  small value is paired with $K{=}9$ Monte Carlo step values that absorb
  V-noise we do not have.
\item Standardized advantage clip $\pm 3\sigma$ retained as an outlier
  safety net (the upstream-faithful no-clip choice survives at \vineppo{}
  but causes \caspo{} to occasionally take large policy steps that
  destabilize online $V_\phi$).
\end{itemize}
\grpo{} and \vineppo{} are unaffected by these knobs at runtime since
they have no learned baseline to drift; we hold the same values across
methods for stack consistency.

\paragraph{$V_\phi$ retrain after verifier and tokenization changes.}
$V_\phi$'s offline training data was collected before two upstream-aligned
patches (Minerva-style answer-extraction cascade in the verifier;
explicit BOS prepending in the rollout engine). At RL time, this leaves
$V_\phi$ scoring \emph{prefixes that begin with a different first token}
(\texttt{<s>} vs.~\texttt{[}) against \emph{outcomes from a verifier that
accepts a broader set of answer formats}. The offline-trained $V_\phi$
reaches step-1 \texttt{v\_acc}$=0.32$ against runtime, well below its
offline-train accuracy of $\sim 0.96$. Re-running the data-collection
pipeline with the live verifier and rollout engine, then retraining
$V_\phi$ for three epochs at the same paper-faithful $5{\times}10^{-7}$
offline learning rate, lifts step-1 \texttt{v\_acc} to $0.74$ at
runtime --- a $2.3\times$ recovery from the stale-label/shifted-prompt
mismatch. We use this retrained $V_\phi$ for all reported \caspo{} runs.

\begin{figure}[t]
\centering
\begin{tikzpicture}[
    font=\footnotesize,
    box/.style={rectangle, rounded corners=2pt, draw, thick, align=center, inner sep=3pt, minimum height=8mm},
    pol/.style={box, fill=black!10, minimum width=18mm},
    vllm/.style={box, fill=black!3, minimum width=18mm},
    ipc/.style={<->, >=Stealth, thick, dashed},
    arr/.style={->, >=Stealth, thick},
    hdr/.style={font=\footnotesize\bfseries},
]
% =========================================================
% Top row: 1B per-method colocated layout
% =========================================================
\node[hdr] (hdr1) at (0,0) {(a) 1B per method (single H100, four methods $\to$ four GPUs)};
\node[pol, below=3mm of hdr1] (p1) {Trainer\\(policy + AdamW)\\fp32 master};
\node[vllm, right=14mm of p1] (v1) {vLLM rollout\\KV $u{=}0.30$};
\draw[ipc] (p1) -- node[above, font=\scriptsize]{CUDA IPC} (v1);

\node[font=\scriptsize, below=2mm of p1, align=center] (cap1)
    {fp32 master via\\\texttt{autocast(cache\_enabled=False)}};
\node[font=\scriptsize, below=2mm of v1, align=center] (capv1)
    {single-rank IPC\\handle exchange};

% =========================================================
% Bottom row: 7B FSDP + per-rank vLLM
% =========================================================
\node[hdr, below=12mm of cap1.south west, anchor=west] (hdr7)
    {(b) 7B FSDP=4 (or 8 for \vineppo{}) + rank-local colocated vLLM};

% Four trainer ranks (top sub-row of the 7B section)
\node[pol, below=4mm of hdr7.south west, anchor=north west] (p7a) {rank 0\\policy shard};
\node[pol, right=2mm of p7a] (p7b) {rank 1\\shard};
\node[pol, right=2mm of p7b] (p7c) {rank 2\\shard};
\node[pol, right=2mm of p7c] (p7d) {rank 3\\shard};

% FSDP all-gather brace above
\draw[<->, thick] (p7a.north) ++(0,3mm) -- ++(0,2mm) -| node[pos=0.25, above, font=\scriptsize]{FSDP \texttt{summon\_full\_params}} (p7d.north);

% Four vLLM engines (bottom sub-row)
\node[vllm, below=8mm of p7a] (v7a) {vLLM\\engine 0};
\node[vllm, below=8mm of p7b] (v7b) {vLLM\\engine 1};
\node[vllm, below=8mm of p7c] (v7c) {vLLM\\engine 2};
\node[vllm, below=8mm of p7d] (v7d) {vLLM\\engine 3};

% Per-rank IPC arrows (vertical)
\draw[ipc] (p7a) -- (v7a);
\draw[ipc] (p7b) -- (v7b);
\draw[ipc] (p7c) -- (v7c);
\draw[ipc] (p7d) -- (v7d);

% Caption strip below the 7B layout
\node[font=\scriptsize, below=2mm of v7b.south, align=center, xshift=8mm] (cap7)
    {streaming IPC push, chunks of 8 with bf16 cast \quad{}\textbar{}\quad{}
     AdamW state CPU-offloaded across the \texttt{summon} window};
\end{tikzpicture}
\caption{Trainer/rollout topology. \emph{(a)} Each 1B run colocates trainer
and vLLM on a single H100 (four methods $\to$ four GPUs of the same node);
fp32 master is maintained via \texttt{autocast(cache\_enabled=False)} so
AdamW covers all parameters, and the IPC handle exchange between the
single-rank trainer and its vLLM engine is direct. \emph{(b)} Each 7B run
uses FSDP=4 (FSDP=8 for \vineppo{}) with one rank-local vLLM engine per
rank. Trainer-to-vLLM weight sync streams parameters in chunks of 8 with
per-tensor bf16 cast inside an
\texttt{FSDP.summon\_full\_params(writeback=False)} block; AdamW state is
offloaded to pinned host memory across the summon window so the
$\sim$28~GiB unsharded fp32 materialization fits alongside the vLLM KV
cache reservation.}
\label{fig:sysarch}
\end{figure}

\section{Results}
\label{sec:results}

This section will report wall-clock and accuracy after the full training and
evaluation runs complete. The intended structure is:

\begin{itemize}
\item Per-method MATH-500 pass@1 and pass@$16$ at \texttt{final}, with
seed-level error bars over a small number of seeds (target: 3 seeds per
method per scale). Methods include \grpo{} ($\mu{=}1$ canonical and
$\mu{=}2$ iso-budget), \ppocritic{}, \vineppo{}, and the four \caspo{}
configurations: \caspo{}-$\Delta p$, \caspo{}-$\Delta\log p$,
\caspo{}-$\Delta p$+refresh, and \caspo{}-$\Delta\log p$+refresh.
\item Per-method full-suite (MATH-500, GSM8K, OlympiadBench) pass@$1$
greedy and pass@$16$ at \texttt{final}.
\item Training curves on MATH-500 sample evaluation at intermediate
checkpoints (every 50 outer steps) plus \texttt{final}, summarizing
sample efficiency.
\item Per-formulation comparison of $\Delta p$ vs.\ $\Delta\log p$ on
MATH-500 pass@$16$, mapping the empirical winner to the saturation
regime predicted in Section~\ref{sec:transforms} and visualized in
Figure~\ref{fig:transforms}.
\item \caspo{} vs.\ \caspo{}+refresh: with the no-refresh default
isolating the value of the Phase-1 PRM and \caspo{}+refresh
quantifying the additional contribution of the iterated PRM-refresh
path (Section~\ref{sec:refresh}).
\item \ppocritic{} vs.\ \caspo{}: contrast a per-token learned scalar
baseline against a per-step learned prefix value, both under the same
PPO objective and the same global PPO minibatch.
\item Wall-clock summary: per-method step time, per-method 1000-step ETA,
and the implied GPU-hour cost of the four-method comparison, with the
breakdown by phase (rollout, old-logprob rescore, value/critic forward,
ref-logprob precompute, policy update, IPC sync).
\end{itemize}

We report all numbers under the same training contract: identical SFT
initialization, identical prompt set, identical verifier, identical
group/response budget, identical PPO clipped objective, identical KL
coefficient, identical advantage standardization, identical max sequence
length, and identical evaluation sampling parameters. The only intended
difference between rows is the source of the per-step value $V(s_t)$ in
\eqref{eq:adv}: zero with a per-prompt group baseline (\grpo{}), a learned
per-token scalar value $V_\psi(s,t)$ (\ppocritic{}), Monte Carlo
continuations (\vineppo{}), or a learned prefix value $V_\phi(s_t)$
(\caspo{}).

\section{Analysis}
\label{sec:analysis}

This section will examine, given the empirical results, what determines when
$V_\phi$ matches \vineppo{}'s prefix-rollout signal and when it does not. The
intended structure is:

\begin{itemize}
\item \emph{Step-credit calibration}: do step advantages from $V_\phi$ rank
correct vs.\ incorrect step continuations consistently with on-policy MC?
\item \emph{PRM decay vs. refresh}: how quickly does the Phase-1
$V_\phi$ lose calibration as the policy drifts (Spearman $\rho$ vs.\
fresh MC labels per outer step), and how much of that decay does the
\caspo{}+refresh schedule recover at each refresh trigger?
\item \emph{Wall-clock per accuracy point}: the practical question \caspo{}
is asked to answer is whether learned prefix values can match \vineppo{}'s
accuracy at a fraction of the wall-clock cost. We report this as
accuracy-versus-GPU-hours for each method.
\item \emph{Step-count effects on \vineppo{}}: \vineppo{}'s wall-clock
scales as $K \cdot \text{steps/response}$, so the marginal cost of step-level
credit grows with response length. We quantify this by binning the test set
into long vs.\ short reasoning regimes and reporting the per-bin
accuracy/wall-clock trade-off.
\end{itemize}

\section{Limitations}
\label{sec:limitations}

\paragraph{Reward-model drift.}
A $V_\phi$ trained on the SFT distribution decays in calibration as
the policy optimizes; the prefixes encountered later in training are
no longer drawn from the SFT marginal, and Spearman $\rho$ vs. fresh
MC labels falls monotonically (Section~\ref{sec:refresh}).
\caspo{}+refresh addresses this by periodically retraining $V_\phi$
from scratch on Monte-Carlo-labeled rollouts of the current policy
(every $k{=}150$ outer steps for $\Delta p$, $k{=}200$ for
$\Delta\log p$). Per-refresh wall-clock is approximately one-tenth
of one \vineppo{} outer-step budget at $K{=}9$, so the method
remains dominated by RL rollout cost rather than refresh overhead.
Neither the no-refresh default nor refresh is a complete defense
against pathological feedback loops (e.g.\ a value head that learns
to predict policy surface form rather than correctness); we report
both to quantify the contribution of the refresh path.

\paragraph{Cross-distribution probe-set evaluation of $V_\phi$.}
Selecting the best refresh recipe by ranking candidate $V_\phi$ checkpoints
on a fixed probe set is sensitive to the response-length cap used when the
probe data was generated. A $V_\phi$ trained on collection budget $L_{\text{tr}}$
sees prefixes spanning the range $[0, L_{\text{tr}}]$, but a probe set
generated at cap $L_{\text{probe}} < L_{\text{tr}}$ contains prefixes only in
$[0, L_{\text{probe}}]$. Differences in Spearman $\rho$ between checkpoints
trained at different $L_{\text{tr}}$ on a single shared probe set therefore
conflate intrinsic credit-assignment quality with the fraction of the
checkpoint's training distribution that the probe never exercises. We
encountered this confound in early refresh-recipe sweeps and recommend
that any future $V_\phi$ recipe selection generate the probe at the
maximum training cap of any candidate, or otherwise restrict the
comparison to recipes whose training caps match the probe cap.

\paragraph{Segmenter quality.}
The LaTeX-aware step splitter is a structural heuristic. Manual spot checks
on Qwen2.5-Math-1.5B and DeepSeekMath-7B generations show coherent
boundaries for prose, equations, factorization, and aligned derivations,
but generations that continue past a final boxed answer have those trailing
fragments treated as additional steps; a post-answer truncation pass is
left as future work.

\paragraph{Apples-to-apples scope.}
Our comparison fixes the SFT initialization, prompt set, verifier, sampling
parameters, KL constraint, and advantage standardization. It does not
exhaustively explore method-specific tuning (e.g.\ different $K$ for
\vineppo{}, different $\beta/m$ for \caspo{}, larger $G$ for \grpo{}); we
report the upstream-paper values where available and otherwise the values
that minimize wall-clock without regressing pass@1 on a small validation set.

\paragraph{Scale.}
Results at 1.5B and 7B may not transfer to substantially larger scales. The
7B FSDP+vLLM-IPC stack is novel to this work; whether the same configuration
extrapolates to 70B is an open systems question.

\section{Future Work: Optimizing the Success-Prediction Architecture}
\label{sec:future-work}
Our results adopt \ipvrm{}'s cumulative-log-ratio parameterization for
$V_\phi$ unchanged, after empirically confirming it dominates a
single-sigmoid-head alternative under matched hyperparameters
(Section~\ref{sec:related}). The architectural margin appears to come
from gradient density --- the log-ratio decomposes loss as a sum of
per-token contributions, so every response token in the prefix supplies
a direct gradient signal --- but several design questions remain open
within this regime.

\paragraph{Heads, depths, and ensembles.}
We trained $V_\phi$ with a single learnable backbone whose log-ratio is
read directly. Future work should compare (i) deeper non-linear value
heads on top of the cumulative log-ratio (e.g.\ a 2--3 layer MLP that
maps $V_\phi(s_t)$ to a scalar logit), (ii) Q-style critics that score
each per-token continuation rather than a prefix value, and (iii)
ensembles of $V_\phi$ trained with different seeds or label collections,
both as variance-reduction tools and as honest-uncertainty signals for
the step-TD advantage.

\paragraph{Differential learning-rate schedules.}
Both architectures we report use a single LR over the entire backbone.
Standard practice for fresh value heads is a 10--100$\times$ higher LR
than the pretrained backbone with a brief warmup; this could close some
of the gap on the sigmoid-head baseline and may also help the log-ratio
PRM at smaller training-data sizes. We did not sweep this axis.

\paragraph{Training-data form.}
We use Math-Shepherd-style continuous-target BCE on per-prefix Monte
Carlo $\widehat{p}$ with $J=8$ continuations and $S=5$ step prefixes
per response. Promising variants include: increasing $J$ for tighter
$\widehat{p}$ variance at fixed compute through bucketed batching;
calibration-aware losses (focal BCE, temperature-scaled targets) when
the underlying success-rate distribution is bimodal; and contrastive
losses pairing within-prompt prefixes of different success rates.

\paragraph{Inference-time refinements.}
At inference, $V_\phi$ is queried in a single forward pass. Test-time
ensembling (multiple seeds, MC-dropout, temperature scaling on
$V_\phi/\beta$) and online calibration on a small held-out probe set
during RL could improve both the calibration and the action of the
step-TD advantage on \caspo{} performance.

\paragraph{Held-out probe protocol.}
The empirical comparison reported here uses a prompt-level held-out
split (\texttt{--split\_by\_prompt} or \texttt{--held\_out\_data}); the
prefix-level row shuffle that the trainer still supports as a legacy
default exhibits same-prompt leakage and should not be used for any
load-bearing claims about $V_\phi$ generalization. Future PRM
ablations should default to held-out prompts and quantify the gap
between in-distribution and OOD performance as a single reported
column.

\paragraph{Minimum RL training-set size for benchmark transfer.}
We train RL on the full 1209-prompt \texttt{dsr\_sub} set and observe
strong generalization to MATH-500, GSM8K, OlympiadBench, AMC, AIME, and
\texttt{college\_math} despite the training-prompt count being orders
of magnitude smaller than typical instruction-tuning corpora. Concretely
the entire RL-training distribution fits in a fixed 1209-prompt subset
of \texttt{DeepScaleR}, yet pass@1 lifts on out-of-distribution
benchmarks are competitive with much larger PPO recipes
(Section~\ref{sec:results}). This raises a sharp practical question:
\emph{how small can the RL prompt set be while still yielding the same
benchmark transfer?} A 300-prompt RL set, for instance, would cut both
the PRM training wall-clock and the RL rollout wall-clock by roughly
$4{\times}$ relative to the 1209-prompt default --- both phases scale
roughly linearly with prompt count at fixed $K$, $J$, and rollouts
per RL step --- and the iterated-refresh schedule (\S\ref{sec:refresh})
would amortize even further if the per-iteration PRM is cheaper.
Promising next experiments: (i) sweep prompt-set sizes
$N \in \{100, 300, 600, 1209\}$ at fixed RL budget and report the
benchmark-transfer Pareto curve; (ii) compare a fixed random subset to
a curated difficulty-balanced subset (since success-rate distributions
shift with the prompt mix, per Section~\ref{sec:vphi}); and (iii) test
whether a small high-quality RL prompt set combined with a high-quality
\emph{PRM} (trained on a separate larger collection) can match the full
1209 RL recipe at a fraction of the wall-clock. The structural
advantage of \caspo{} over methods that require on-policy MC
continuations (\vineppo{}) is exactly that the PRM is decoupled from
the RL prompt set, so the two budgets can be chosen independently.

\section{Conclusion}
\label{sec:conclusion}

We described \caspo{}, a step-level credit assignment method for
reasoning RL that replaces \vineppo{}'s on-policy Monte Carlo prefix
rollouts with a learned process reward model trained on Monte Carlo
step labels to predict $P(\text{success} \mid \text{prefix})$, and a
periodic refresh of that model on current-policy rollouts to control
distribution drift. \caspo{} fits cleanly into the same PPO clipped
objective, the same KL constraint, and the same step-TD advantage
construction used by \vineppo{}, isolating the credit-signal change
from confounders. By unifying
\grpo{}, \ppocritic{}, \vineppo{}, and \caspo{} under one trainer, one
verifier, and one rollout engine, we make it possible to compare these four
methods under identical conditions on Qwen2.5-Math-1.5B with
\texttt{dsr\_sub} as the primary site and DeepSeekMath-7B with
MATH-lighteval as a paper-faithful 7B reference. Empirical results in
Section~\ref{sec:results} will quantify whether the learned value model
recovers \vineppo{}'s reasoning-credit signal at a fraction of its
generation cost, or whether step-level Monte Carlo remains uniquely
advantageous, and whether iterated refresh of $V_\phi$ from current-policy
rollouts closes any residual gap.

\bibliographystyle{plainnat}
\bibliography{references}

\end{document}
